\documentclass[a4paper]{article}
\usepackage[utf8]{inputenc}


%=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-
% PREAMBLE
%=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Important styling notes
%%
% For now, to include img.jpg in img/path/to/img.jpg, just use:
% path/to/img.jpg - for details see style.tex
\input{style.tex}




\begin{document}
%=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-
% GLOBAL STYLES (DOCUMENT SCOPE)
%=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-
% caption: Figure 1 -> <bold> Fig. 1 </bold>
\captionsetup[figure]{labelfont={bf},labelformat={default},labelsep=period,name={Fig.}}


%=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-
% TITLE PAGE
%=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-
\input{title.tex}
%\maketitle



%=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-
% MAIN DOCUMENT
%=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-
\newpage
\tableofcontents
\newpage



%------------------------------ New section ------------------------------%
\section{Path priority and executing binaries together}



\subsection{Motivation}
I was browsing the \href{https://wiki.archlinux.org/index.php/Spotify#blockify}{Arch wiki} on how to automatically launch \texttt{blockify} when \texttt{spotify} is launched and how to kill it in the same manner.

The wiki instructs to place the following script at \texttt{/usr/local/bin/spotify}:
\begin{lstlisting}[language=bash]
#!/bin/sh

spotify=/usr/bin/spotify

if [[ -x $spotify && -x /usr/bin/blockify ]];
then
  blockify &
  block_pid=$!
  $spotify
  trap "kill -9 $block_pid" SIGINT SIGTERM EXIT
fi

\end{lstlisting}

First, let's try to recreate something similar with some simple dummy executables.

\subsection{How path priority works}

By path priority, I refer to the sub-paths in the system's \texttt{PATH} environmental variable. The latter variable tells a Unix system which directories to search when executing a program (command). The directories in the path are separated by \texttt{:}. The latter variable looks like:
\begin{verbatim}
PATH="dir1:dir2:...:dirN"	
\end{verbatim}
When executing a command, its executable is first searched in \texttt{dir1}, then \texttt{dir2}, then $\ldots$ \texttt{dirN} so the priority is:
\[
	Pr(dir_1) > Pr(dir_2) > \ldots > Pr(dir_N)
\]
A typical \texttt{PATH} in a Unix system contains the following:
\begin{verbatim}
/usr/local/sbin:/usr/local/bin:/usr/bin:...
\end{verbatim}
, which shows what the priority is. 

Going back to the blockify script, since \texttt{/usr/local/bin/spotify} has greater priority than \texttt{/usr/bin/spotify}, the script gets executed instead of plain \texttt{spotify}. The script executes blockify and stores its PID (\texttt{\$!}) in a variable


\subsection{Traps; waiting for signal to execute action}

The next goal of the script is to kill \texttt{blockify} when \texttt{spotify} terminates, achieved by the line:
\begin{verbatim}
trap "kill -9 $block_pid" SIGINT SIGTERM EXIT
\end{verbatim}
In Unix, traps are used to activate handlers when a particular signal is received. The syntax to set up a trap is the following:
\begin{verbatim}
trap <"handler command"> <signal list>
\end{verbatim}
In this case, the handler is to kill blockify, i.e. \texttt{kill -9 \$block\_id}. We want the handler to be reached when signals \texttt{SIGINT, SIGTERM, EXIT} are received, i.e. when spotify terminates. Signals \texttt{SIGINT}, \texttt{SIGTERM} are received when a command terminates the program and \texttt{EXIT} is a pseudo-signal received when the program exits. That's it.


\subsubsection{Traps; a simple example}

Let's say we have a script keeps writing random numbers from 0 to 99 to a file and is meant to remove the file when it exits, either manually or by itself.

\begin{lstlisting}[language=bash]
#!/bin/bash

output_file=/tmp/output.txt
> $output_file 

# --- trap handler
cleanup() {
	echo "Exit received. Cleaning up..."
	rm $output_file 
}
# ---

trap "cleanup" SIGINT SIGTERM EXIT


#--- main work 
for i in `seq 1 10`; do
	echo $[ $RANDOM \% 100 ] >> $output\_file 
	sleep 1
done
#---
\end{lstlisting}
When the user kills the program by \texttt{\^C} or when the program exits, it calls the handler \texttt{cleaup()} and removes the output file.


%=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-
% References
%=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-
\newpage
%\printbibliography



%=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-
% Appendices
%=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-=-
%\newpage
%\appendix

%\section{Appendices}

% ------------------------ New appendix ------------------------ %
%\newpage
%\subsection{Appendix Example}
%\label{app:my_cool_appendix}




\end{document}
